Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-12-01 analysis/paper/paper.qmd
added comments for GP's tasks to revise the manuscript

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA). We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias. Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more due to guided variation. Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group, and then modified the shape according to their specific circumstances.

@@

To analyse change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition. Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to analyse them using these three phases to facilitate observations of change over time.

@@

@@

threshold_value <- 25

@@

To measure the variation among the measurements of the stemmed points, we computed Coefficients of Variation (CV) with confidence intervals. The CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarized CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e. CV for planned production = 1.7) and about 60 (i.e. CV for random production = 57.7) and the values over the upper limit likely reflect  attempts to intentionally differentiate the products. Among the Great Basin projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea compared to the Great Basin, and the median of the CV range (30.5) for the Korean lithics, we selected `r threshold_value` as the threshold to decide between guided variation and indirect bias. We premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias. <!-- I STILL THIS WE NEED A MORE COMPELLING JUSTIFICATION FOR THE THRESHOLD VALUE, BASED ON SOME THEORY IN THE LITERATURE. I EXPECT REVIEWERS WILL GET STUCK ON THIS AND DEMAND MORE DETAILS. THE CHOICE OF VALUE HAS A HUGE EFFECT ON OUR INTERPRETATIONS BELOW -->

@@

Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artefacts is higher. Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was influenced by indirect bias.

```{r}
#| label: fig-plot-cv-all-attributes
#| fig-cap: A. Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value. The dots represent individual artefacts. B. The points and blue colored numbers represent CV values for each attribute measured on the stemmed points. 95% confidence intervals for the CVs were computed using Sharma and Krishna's method. The vertical lines indicate the range of confidence intervals.

@@

```{r}
#| label: fig-plot-cv-over-time
#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   4.5
#| fig-height:  4

@@

cv_by_stage_df  <-
  df_sitename %>%
  dplyr::select(-sitename,
         -SPstage.Raw_material) %>%
  group_by(SPstage.Stage) %>%
  add_tally() %>%
  dplyr::filter(SPstage.Stage > 1) %>%
  dplyr::select(-n) %>%
  pivot_longer(-SPstage.Stage,
               names_to = "variable",
               values_to = "value") %>%
  group_by(SPstage.Stage, variable) %>%
  dplyr::summarise(cv_by_site =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

@@

One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-over-time). These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making. The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS). SYG1_2 and SYG6_2 are from Phase 2 (shown on the left side of the figure) and SYG1_3 and YS_3 are from Phase 3 (shown on the right side of the figure). Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly below `r threshold_value`, which represent transmissions dominated by indirect bias. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-over-time. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of indirect bias, rather than substantial changes in the type of bias dominating cultural transmission.

```{r}
#| label: fig-plot-cv-four-assemblage
#| fig-cap: CV values of four attributes from assemblages with more than 5 stemmed points. Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
#| fig-width:   4.5
#| fig-height:  4

@@

```{r}
#| label: fig-plot-cv-by-raw-material
#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   8.5
#| fig-height:  4

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Law Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() 
```

@@

<!-- GP to write up the results in detail, that you see in the plot made by the code below. Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology" -->

```{r}
#| label: fig-corr-analysis
#| fig-cap: Correlation analysis of point attributes, calculated using Pear
#| fig-width:   7
#| fig-height:  4

@@

p3_corr_plot <-  
  rplot(df_sitename_corr_p3) +
   theme(legend.position = c(0.8, 0.8))

@@

# compare p2 and p3 correlations
compare_p2_and_p3_correlations_plot <- 
list(`phase 2` = df_sitename_corr_p2_long,
     `phase 3` = df_sitename_corr_p3_long) %>% 
  bind_rows(.id = "phase") %>% 
  ggplot() +
  aes(phase,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 4,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient")

library(cowplot)
plot_grid(
          plot_grid(p2_corr_plot, 
                    p3_corr_plot,
                    ncol = 1,
                    labels = c("A", "B")),
          compare_p2_and_p3_correlations_plot,
          ncol = 2,
          labels = c("", "C"),
          rel_heights = c(1, 1.5))

@@

The CV values for all stemmed point attributes are mostly over 25, except for maximum length, which is just below. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified signed-likelihood ratio test (SLRT) to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

However, similar to the CV variation for the Phase 2,  This combination model describes that there were social connections between groups that share the tool production technique. Individuals or groups learned the major technology of stemmed points by copying a model from another individual or group (indirect bias). Then, those individuals or groups adjusted their technique and shape of the stem depending on their needs and material availability. Our results recalls -@lee2013current's model but his model explains combination of the *in situ* tradition for flake tools and introduction of blade-associate toolkit including stemmed points. On the other hand, we saw the combination of the *in situ* tradition and introduced stemmed point technology within the single artifact.

@@

In this research, we examine the dominant mode of cultural transmission that might impact technological innovation in the Korean Late Paleolithic. Following the previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we use two transmission biases, guided variation and indirect bias to scrutinize the transmission process and social contexts. We claim the two scenarios that explains the introduction of new technology: socially isolated groups made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others (indirect bias). We ask three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? do the modes of cultural transmission vary over time? and do the modes of cultural transmission vary over space? To determine transmission biases and answer our research questions, we computed CV and PCA. The higher CV values (i.e. \>=25) and low correlation indicate guided variation while the lower CV while the lower CV values (i.e. \<25) and high correlations represent indirect bias. To offset the small sample issue and increase the credibility of the comparisons, we applied confidence intervals to CV for better estimation.

Our results show that some attributes such as body length, maximum length, mid width, and tang width were transmitted through indirect bias and the other attributes were developed through guided variation. All attributes are positively related and those low variation attributes are more closely associated and highly contribute to the overall shape of the stemmed points. We assume that the guided variation played a more important role at the beginning of the transition based on high CV values of Phase 1. We observe a slight decrease in CV values over time. No clear regional patterns are confirmed. CV values could be depending on raw materials. For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals. Overall, we conclude that combination of the two transmission biases well explains the technological transition during the Late Paleolithic. We anticipate that our theoretical approach is applicable to other studies about technological transition and cultural transmission. We used landmarks to earn the dimensional attributes, which has potential for future research with different dimensions by calculating the distance between other sets of the existing landmarks. In our future research, we will focus on understanding actual connection between the assemblages and examining the social networks for sharing technology between the groups such as trades or physical migration to different locations.

>>>>

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and correlation coefficients. We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias. Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more due to guided variation. Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group, and then modified the shape according to their specific circumstances. <!-- GP to revise and edit the abstract after completing the other tasks below -->

@@

To analyse change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition. Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.

Panel A of @fig-raw-materials-by-phase summarises the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases . We exclude phase 1 from our analyses below because of the small sample size. Shale is the dominant raw material in phase 2 and 3. In phase 3 there is an increase in the proportions of rhyolite and porphyry, driven largely by finds from Wolpyeng and Yongsangdong. Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at sites with more than five stemmed points. Much of the raw material diversity comes from isolated finds, with most sites with more than five points showing high homogeneity in raw materials.

<!-- BM: I have left many captions with only xxx. GP please fill them out with detailed and specific captions -->

```{r}
#| label: fig-raw-materials-by-phase
#| fig-cap: "A: xxx. B: Frequencies of stemmed points by site and raw material, for sites with five or more stemmed points "
#| fig-height: 4

df_full_sitename_fct <- 
df_full_sitename %>% 
  mutate(SPstage.Raw_material = factor(SPstage.Raw_material))

#Create a custom color scale
library(RColorBrewer)
myColors <- 
  brewer.pal(length(unique(df_full_sitename_fct$SPstage.Raw_material)),
             "RdBu")
names(myColors) <- levels(df_full_sitename_fct$SPstage.Raw_material)
colScale <- scale_fill_manual(name = "Raw material",
                                values = myColors)

p1 <- 
ggplot(df_full_sitename_fct) +
  aes(SPstage.Stage,
      fill = SPstage.Raw_material) +
  geom_bar() +
  theme_bw() +
  labs(fill = "Raw material",
       x = "Chronological Phase") +
  theme(legend.position = "top",
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        legend.key.height = unit(0.5, 'cm'), #change legend key height
        legend.key.width = unit(0.5, 'cm'), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)
        ) +
colScale

p2 <- 
  df_full_sitename_fct %>%
  mutate(SPstage.Stage = paste0("Phase ", SPstage.Stage)) %>% 
  group_by(full_sitename) %>% 
  filter(n() >= 5) %>% 
    ggplot() +
    aes(full_sitename,
        fill = SPstage.Raw_material) +
    geom_bar() +
    facet_wrap(~ SPstage.Stage) +
  theme_bw () +
  labs(fill = "Raw material",
       x = "Site") +
 theme(legend.position = "none",
       axis.text.x = element_text(size = 4)  ) +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
   colScale

library(cowplot)
plot_grid(p1, p2,
          ncol = 1,
          rel_heights = c(1.5, 0.75),
          labels = c("A", "B"))
```

@@

<!-- BM: I think these tables should appear before the figures above, what do you think? What do you think a typical reader is expecting? -->

@@

# to ensure we don't have any copy-paste errors
threshold_value <- 25
```

@@

To measure the variation among the measurements of the stemmed points, we computed Coefficients of Variation (CV) with confidence intervals. The CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarized CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e. CV for planned production = 1.7) and about 60 (i.e. CV for random production = 57.7) and the values over the upper limit likely reflect  attempts to intentionally differentiate the products. Among the Great Basin projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea compared to the Great Basin, and the median of the CV range (30.5) for the Korean lithics, we selected `r threshold_value` as the threshold to decide between guided variation and indirect bias. We premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias. <!-- BM: I STILL THIS WE NEED A MORE COMPELLING AND STRONGER JUSTIFICATION FOR THE THRESHOLD VALUE, BASED ON SOME THEORY IN THE LITERATURE. PLEASE ADD SOMETHING. I EXPECT REVIEWERS WILL GET STUCK ON THIS AND DEMAND MORE DETAILS. THE CHOICE OF VALUE HAS A HUGE EFFECT ON OUR INTERPRETATIONS BELOW -->

@@

<!-- BM: overall I think the CV values are really not that interesting or informative. They are difficult to interpret because they all cluster around 15-30, which seems to me to be the midpoint between 'highly standardised' (3-5) and 'highly different' (40-50). So I think we will struggle to convince any readers that we have any strong signals of cultural transmission biases in the CV data. Probably we can satisfy most readers that guided variation is the main bias in these data. That's my perspective after reading this paper several times, and re-reading several related papers by Bettiner and Eerkens and Lipo, and others  -->

Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artefacts is higher. Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was more influenced by indirect bias than the others.

```{r}
#| label: fig-plot-cv-all-attributes
#| fig-cap: A. Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value. The dots represent individual artefacts. B. The points and blue colored numbers represent CV values for each attribute measured on the stemmed points. 95% confidence intervals for the CVs were computed using Sharma and Krishna's method. The vertical lines indicate the confidence intervals.

@@

```{r}
#| label: fig-plot-cv-over-time
#| fig-cap: CV values for the second and third chronological phases in the Korean Late Paleolithic period. The points represent CV values for each attribute. The vertical lines indicate the confidence intervals.
#| fig-width:   4.5
#| fig-height:  4

@@

cv_by_stage_df  <-
  df_sitename %>%
  dplyr::select(-sitename,
         -SPstage.Raw_material) %>%
  group_by(SPstage.Stage) %>%
  add_tally() %>%
  dplyr::filter(n >= 5) %>%
  dplyr::select(-n) %>%
  pivot_longer(-SPstage.Stage,
               names_to = "variable",
               values_to = "value") %>%
  group_by(SPstage.Stage, variable) %>%
  dplyr::summarise(cv_by_site =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

@@

One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-four-assemblage). These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making. The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS). SYG1_2 and SYG6_2 are from Phase 2 (shown on the left side of the figure) and SYG1_3 and YS_3 are from Phase 3 (shown on the right side of the figure). Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.

```{r}
#| label: fig-plot-cv-four-assemblage
#| fig-cap: CV values of four attributes from assemblages with more than 5 stemmed points. Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3. The points represent CV values for each attribute. The vertical lines indicate the confidence interval.
#| fig-width:   4.5
#| fig-height:  4

@@

```{r}
#| label: fig-plot-cv-by-raw-material
#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the confidence intervals.
#| fig-width:   8.5
#| fig-height:  4

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Raw Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() 
```

@@

<!-- BM: You'll see I've completely removed the PCA content. I did some more reading into what PCA calculates, and it is not correlation coefficients of the variables with each other, but with the Principal Components that it calculates. So that's a very different kind of correlation, compared to what we see in the previous literature on transmission biases. I suppose we could use it and interpret it, if you want to investigate that further. I could not think of anything straightforward over the several days I have been editing this. I saw that you did not give them the same treatment as the CV values with exploring variation by raw material and by site, so I guess you were not very interested in the PCA results. My perspective is that we should look for the quickest and simplest way to complete this paper. So I have replaced the PCA with the same kind of correlation analysis that we see elsewhere so we can use the previously published papers to guide our interpretation of our results -->

<!-- BM: GP to write up the correlation coefficients results in detail, according to what you see in the plot made by the code below. I've already done all the analysis here. Your task is to describe it here, and then contextualise and interpret it in the next section. Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology". Aim for a style of writing that is detailed, direct, specific, short sentences.  -->

```{r}
#| label: fig-corr-analysis
#| fig-cap: "A: xxx. B. xxx. C. xxx"
#| fig-width:   7
#| fig-height:  9

# BM: I added this section of analysis to replace the PCA sections.

@@

p3_corr_plot <-  
  rplot(df_sitename_corr_p3) +
   theme(legend.position = c(0.9, 0.9),
         legend.key.size = unit(0.3, 'cm'), #change legend key size
        legend.key.height = unit(0.3, 'cm'), #change legend key height
        legend.key.width = unit(0.3, 'cm'), #change legend key width
        legend.title = element_text(size=6), #change legend title font size
        legend.text = element_text(size=8))

@@

# compare p2 and p3 correlations
compare_p2_and_p3_correlations_plot <- 
list(`phase 2` = df_sitename_corr_p2_long,
     `phase 3` = df_sitename_corr_p3_long) %>% 
  bind_rows(.id = "phase") %>% 
  ggplot() +
  aes(phase,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 2,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient")

# correlation by raw material
df_sitename_corr_raw_material <- 
  df_sitename %>%
  group_by(SPstage.Raw_material) %>% 
  filter(n() >= 3) %>% 
  dplyr::select(-sitename, 
                -SPstage.Stage) %>% 
  nest(-SPstage.Raw_material) %>% 
  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
                                   shave() %>% 
                                  pivot_longer(-term))) %>% 
  unnest(corrr) %>% 
  filter()

correlations_by_raw_material <- 
  ggplot(df_sitename_corr_raw_material) +
  aes(SPstage.Raw_material,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 2,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient",
       x = "Raw material")

# correlation by site
df_sitename_corr_site <- 
  df_sitename %>%
  group_by(sitename) %>% 
  filter(n() >= 5) %>% 
  dplyr::select(-sitename, 
                -SPstage.Stage) %>% 
  nest(-sitename) %>% 
  mutate(corrr = purrr::map(data, ~correlate(.x) %>%  
                                   shave() %>% 
                                  pivot_longer(-term))) %>% 
  unnest(corrr) %>% 
  filter()

correlations_by_site <- 
  ggplot(df_sitename_corr_site) +
  aes(sitename,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 2,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient",
       x = "Site")

library(cowplot)
plot_grid(plot_grid(
          plot_grid(p2_corr_plot, 
                    p3_corr_plot,
                    ncol = 1,
                    labels = c("A", "B")),
          compare_p2_and_p3_correlations_plot,
          ncol = 2,
          labels = c("", "C"),
          rel_heights = c(1, 1.5)),
          correlations_by_raw_material,
          correlations_by_site,
          ncol = 1,
          labels = c("", "D", "E"))

@@

The CV values for all stemmed point attributes are mostly over 25, except for maximum length, which is just below. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (SLRT) <!-- GP to add a citation for this test here --> to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

<!-- BM: GP to write up the discussion of the correlation analysis in detail. Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology". Aim to improve your writing, make it better than your usual style. -->

@@

In this research, we investigated the social context of technological innovation in the Korean Late Paleolithic. Following previous studies of applying cultural transmission to the introduction of bow and arrow technology in the Great Basin [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we investigated two transmission biases, guided variation (trial and error) and indirect bias (copying a model). We proposed  two scenarios for explaining the introduction of new technology: socially isolated groups that made stemmed points through trial and error (guided variation) or socially connected groups whose knowledge of stemmed points derived from copying others that they regularly came in contact with (indirect bias). We asked three questions: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? And do the modes of cultural transmission vary over space?

We conclude that the dominant mode of cultural transmission for the earliest stemmed points was guided variation (trial and error). This is indicated by relatively high CV values of around and above 25 and mostly low correlation coefficients of less than 0.5. A social context that favoured trial and error for artefact making is consistent with Seong's _in situ_ model for the appearance of stemmed points in Korea. This model proposes that people developed stemmed points by experimenting with existing blade technologies. Our results support experimentation and trial and error as important processes in the cultural transmission of this artefact technology.

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values at 0.5 and higher in chronological phase 3, after 25 ka. This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. This may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model. We might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure <!-- GP to find something relevant by Fitzhugh to cite here -->. With more frequent contacts and connections with members of the social network, there may have been more opportunities to learn point-making directly from socially successful individuals.

In comparing sites with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive forms of cultural transmission of tool-making at different sites. Correlation coefficient values across the sites affirm the overall chronological trend indicating indirect bias in phase three. Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Wolpyeng and Yonsangdong that date to this period. One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favoured an expanded range for searching for raw materials, leading to a higher diversity. Similarly, higher mobility may have resulted in more frequent contacts and connections with members of social networks. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.

<!-- BM: GP to check where we can cite our QSR paper in the conclusion. Does what we say here agree with what we said in the QSR paper? GP to edit the conclusion here to show good harmony with our previous paper -->.
